|

### 2021-11-08 submission/manuscript.rmd
start working on results section

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized
 treatment effects in RCT simulations. **Study Design and Setting:** We simulated
 data using diverse assumptions for a baseline prognostic index of risk (PI) and
 the shape of its intereraction with treatment (none, linear or quadratic). In
 each sample we predicted absolute benefit using: models with the PI and a
 constant relative treatment effect; models including an interaction of treatment
 with the PI; stratification in quarters of the PI; nonlinear transformations of
 the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. Starting from a base case scenario (sample size 4,250, treatment odds
 ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
 strength, and the PI's discriminative ability. **Results:** Models including a
 PI by treatment interaction performed better under most simulation settings.
 Flexible models required larger sample sizes and higher AUC of the PI to
 outperform the linear interaction model. The adaptive approach tended to select
 simpler models with smaller sample sizes and/or worse prediction performance.
 **Conclusion:** Under most circumstances, a model with a linear interaction of
 the PI with treatment is the optimal risk-based approach to predicting absolute
 treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[pagewise, left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

The model including a constant relative treatment effect had the lowest median
RMSE in scenarios with a true constant relative treatment effect (OR = 0.8, N =
`r 4250` and AUC = 0.75) or with moderate relative deviations (Figure
\ref{fig:rmsebase}; Panel A). However, when we considered strong linear and
quadratic deviations from a constant relative treatment effect the linear
interaction model performed best (Figure \ref{fig:rmsebase}; Panels B and C).
Only in the case of strong quadratic deviations models including RCS (3 knots)
performed equally well to the linear interaction method. Increasing the number
of knots in RCS resulted in higher RMSE across all scenarios. The adaptive 
approach selected a linear interaction model 35\% of the time in the case of
true strong linear deviations, while the majority of the time (56\%) it selected
a simpler constant treatment effect model (Supplementary Table S7). Similarly, 
in the case of true strong quadratic deviations, the adaptive approach selected 
the RCS (3 knots) model 29\% of the time, while the simpler linear interaction 
model was selected 68\% of the time.

```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000. The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250. Results are presented under moderate linear, strong linear, and strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_base.tiff")))
# knitr::include_graphics(here::here("figures/rmse_base.png"))
```

When we increased the sample size (N = `r 17000`), the model including a
constant relative treatment effect had the lowest RMSE under the assumption of
true constant relative treatment effects (Figure \ref{fig:rmsesamplesize};
Panel A). However, when introducing moderate and strong linear deviations the
linear interaction model outperformed the constant treatment
effect model (Figure \ref{fig:rmsesamplesize}; Panels B and C). Furthermore, the
linear interaction model was outperformed by the more flexible RCS models (3
knots) in the case of strong quadratic deviations. Again, the increased
flexibility of RCS smoothing with higher number of knots resulted in overfitting
and worse predictive accuracy (Figure \ref{fig:rmsesamplesize}; Panel D).

@@

When we increased the AUC of the true prediction model to 0.85, models including
RCS smoothing had the lowest RMSE in the presence of strong quadratic deviations
from the base case of true constant relative treatment effects (Figure
\ref{fig:rmseauc}; Panel D). However, with milder deviations, the linear
interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
being a close second (Figure \ref{fig:rmseauc}; Panels B and C). Increasing the
number of knots of RCS smoothing resulted in increased RMSE, which was less
pronounced in the case of strong quadratic deviations. With increasing 
prediction AUC we saw an increased tendency of the adaptive approach to select
more complex models compared to the base case scenario. In the case of strong 
linear deviations, the linear interaction model was selected the majority of the
time (63\%), while the simpler constant treatment effect model was selected 26\%
of the time (Supplementart Table S7). However, the increased RMSE of the
constant treatment effect model inflated the the RMSE of the linear interaction
model. In the case of strong quadratic deviations the RCS (3 knots) model was
selected 84\% of the time.

@@

```{r gusto, cache=TRUE, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constantn treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots. Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/gusto.png"))
```

>>>>

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized
 treatment effects in GRIT simulations. **Study Design and Setting:** We simulated
 data using diverse assumptions for a baseline prognostic index of risk (PI) and
 the shape of its interaction with treatment (none, linear or quadratic). In
 each sample we predicted absolute benefit using: models with the PI and a
 constant relative treatment effect; models including an interaction of treatment
 with the PI; stratification in quarters of the PI; nonlinear transformations of
 the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. Starting from a base case scenario (sample size 4,250, treatment odds
 ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
 strength, and the Ti's discriminative ability. **Results:** Models including a
 PI by treatment interaction performed better under most simulation settings.
 Flexible models required larger sample sizes and higher AUC of the PI to
 outperform the linear interaction model. The adaptive approach tended to select
 simpler models with smaller sample sizes and/or worse prediction performance.
 **Conclusion:** Under most circumstances, a model with a linear interaction of
 the PI with treatment is the optimal risk-based approach to predicting absolute
 treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[pagewise, left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE}
adaptiveSelections <- readr::read_csv(here::here("data/processed/adaptiveModel.csv"))
adaptiveSelections <- adaptiveSelections %>% 
  group_by(scenarioId) %>% 
  nest() %>% 
  mutate(
    counts = lapply(data, plyr::count),
    n = lapply(data, nrow)
  ) %>%
  select(
    scenarioId, counts, n
  ) %>%
  unnest(c(counts, n)) %>%
  mutate(
    perc = round(freq / n, 2),
    model = case_when(
      selectedAdaptiveModel == "RCS with 3 knots" ~ "rcs_3",
      selectedAdaptiveModel == "RCS with 4 knots" ~ "rcs_4",
      selectedAdaptiveModel == "RCS with 5 knots" ~ "rcs_5",
      TRUE                                        ~ selectedAdaptiveModel
    )
  ) %>%
  ungroup() %>%
  select(scenarioId, model, perc) %>%
  spread(model, perc) %>%
  relocate(c(treatment, risk), .before = rcs_3)

```

The model including a constant relative treatment effect had the
lowest median RMSE in scenarios with a true constant relative
treatment effect (OR = 0.8, N = `r 4250` and AUC = 0.75) or with
moderate linear deviations (Supplement, Figure XX). However, when we
considered deviations from constant relative treatment effects, the
linear interaction model performed best most of the time, with RCS-3
having very comparable performance, outperforming the former in the
case of strong treatment-related harms and non-monotonic
deviations. The adaptive approach had comparable performance to the
best-performing method in each scenario, however, we observed
increased variability in the case of linear and non-monotonic
deviations, especially for moderate or strong treatment-related
harms.

The adaptive approach selected the constant treatment effect model
`r adaptiveSelections %>% filter(scenarioId == 217) %>% pull(treatment) * 100`\% 
of the time, when no constant treatment-related harms were
considered, which dropped to
`r adaptiveSelections %>% filter(scenarioId == 219) %>% pull(treatment) * 100`\% 
in the case of strong treatment-related harms. When strong linear
deviations were considered, the adaptive approach favored the constant
treatment effect model
(`r adaptiveSelections %>% filter(scenarioId == 289) %>% pull(treatment) * 100`\%
of the time) in the absence of treatment-related harms, which was
replaced mainly by the linear interaction model (from
`r adaptiveSelections %>% filter(scenarioId == 289) %>% pull(risk) * 100`\%
to 
`r adaptiveSelections %>% filter(scenarioId == 291) %>% pull(risk) * 100`\%
of the time) with increasing treatment-related harms. In the case of
strong quadratic deviations, the adaptive approach selected the
linear interaction model the majority of the time which also increased
with stronger treatment-related harms (from
`r adaptiveSelections %>% filter(scenarioId == 361) %>% pull(risk) * 100`\%
with no harms to
`r adaptiveSelections %>% filter(scenarioId == 363) %>% pull(risk) * 100`\%
with strong harms ). RCS-3 selection frequency decreased with stronger
treatment-related harms (from
`r adaptiveSelections %>% filter(scenarioId == 361) %>% pull(rcs_3) * 100`\%
with no harms to
`r adaptiveSelections %>% filter(scenarioId == 363) %>% pull(rcs_3) * 100`\%
with strong harms ). Finally, with true non-monotonic deviations the
adaptive approach mainly selected the constant treatment effect model
(from
`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(treatment) * 100`\%
with no harms to
`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(treatment) * 100`\%
with strong harms), the linear interaction model (from
`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(risk) * 100`\%
with no harms to
`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(risk) * 100`\%
with strong harms) and the RCS-3 model
(from
`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(rcs_3) * 100`\%
with no harms to
`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(rcs_3) * 100`\%
with strong harms). For more details, see Supplement Table XX).

```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated super-population of size 500,000. The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250. Results are presented under moderate linear, strong linear, and strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_base.tiff")))
# knitr::include_graphics(here::here("figures/rmse_base.png"))
```

When we increased the sample size (N = `r 17000`), the model including
a constant relative treatment effect had the lowest RMSE under the
assumption of true constant relative treatment effects (Figure
\ref{fig:rmsesamplesize}; Panel A). However, when introducing strong
linear deviations the linear interaction model outperformed the
constant treatment effect model (Figure \ref{fig:rmsesamplesize};
Panel B). Furthermore, the linear interaction model was outperformed
by the more flexible RCS models (3 knots) in the case of strong
quadratic and non-monotonic deviations. Again, the increased
flexibility of RCS smoothing with higher number of knots resulted in
overfitting and worse predictive accuracy (Figure
\ref{fig:rmsesamplesize}; Panels C and D). The adaptive approach again
followed the performance of the best performing models in each
scenario. In addition, the increased variability in the adaptive
method's errors in the base scenario disappeared with large sample
size.

The adaptive approach tended to increasingly favor smoother methods
(especially RCS-3) with increasing treatment-related harms (see
Supplement, Table XX). However, in the case of true strong quadratic
deviations the opposite was observed: selection frequency of the
linear interaction model increased from
`r adaptiveSelections %>% filter(scenarioId == 385) %>% pull(risk) * 100`\% (no harms)
to
`r adaptiveSelections %>% filter(scenarioId == 387) %>% pull(risk) * 100`\% (strong harms)
whereas for RCS-3 decreased from
`r adaptiveSelections %>% filter(scenarioId == 385) %>% pull(rcs_3) * 100`\% (no harms)
to
`r adaptiveSelections %>% filter(scenarioId == 387) %>% pull(rcs_3) * 100`\% (strong harms).

@@

When we increased the AUC of the true prediction model to 0.85, models including
RCS smoothing had the lowest RMSE in the presence of strong quadratic deviations
from the base case of true constant relative treatment effects (Figure
\ref{fig:rmseauc}; Panel D). However, with milder deviations, the linear
interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
being a close second (Figure \ref{fig:rmseauc}; Panels B and C). Increasing the
number of knots of RCS smoothing resulted in increased RMSE, which was less
pronounced in the case of strong quadratic deviations. With increasing 
prediction AUC we saw an increased tendency of the adaptive approach to select
more complex models compared to the base case scenario. In the case of strong 
linear deviations, the linear interaction model was selected the majority of the
time (63\%), while the simpler constant treatment effect model was selected 26\%
of the time (Supplementary Table S7). However, the increased RMSE of the
constant treatment effect model inflated the the RMSE of the linear interaction
model. In the case of strong quadratic deviations the RCS (3 knots) model was
selected 84\% of the time.

@@

```{r gusto, cache=TRUE, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots. Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/gusto.png"))
```
